Spatial Reasoning and Data Displays

### 2015-06-26 Paper/InfoVis2015Revision/ResponsesToReviewers.Rmd

We would like to thank the reviewers for the detailed comments. We hope to have addressed all of the questions.

We are detailing our response below:

# Major Concerns
1. **Are all the experimental design choices properly motivated?**    
**(refer to the first concern of R1).**    
*We have extended the discussion of the choice of cognitive tests; the purpose of each of the cognitive tests is described in section 2.2. In summary, the paper-folding, card rotation, and figure classification tasks are part of the kit of factor-referenced cognitive tests (and are thus extremely well-studied); additionally, we included a visual searching task (VST) because of its similarity of form to statistical lineups and its' compatibility with the written form of the other cognitive tests and lineup tests.*    
*As mentioned in the paper, the paper folding and card rotation tests have been used in many previous studies of statistical graphics and data displays. The figure-classification task examines inductive reasoning, which is another component of the lineup task (determining the dimension on which the lineup components differ, then applying that to answer the question successfully).*
*Regarding our hypotheses about the relationship of the visuospatial tests and lineups: we extended the discussion in the introduction; in summary we hypothesize that visual search is the basis for any lineup evaluation. We are further intrested in whether the spatial component (measured in 2d by the card rotation test or in 3d by the paper folding test) dominates lineup performance or inductive reasoning (as measured by the figure classification test).*

2. **Are the experimental procedure sufficiently detailed?**    
**(refer to the second concern of R1 and )**

@@

*While there are differences in the correlation of scores on lineup sets 1, 2, and 3 with the cognitive tests, this does not provide any additional information about whether designs in lineup set 1 are superior to those in lineup set 2 and lineup set 3, and in fact, the sets of lineups are not comparable in this way. Each set of lineups was taken from an experiment which did aim to rank designs by effectiveness, but in this study we are instead examining individual differences in cognitive abilities and lineup performance.*
7. **Does the revision resolve a potential concern of confounding factors**    
**(refer to the conern of R3)**    
*The issue of confounding factors is related to the issue raised in #6, that is, that we have not discussed visualization design in this paper. As explained above, our goal was not to compare types of visualizations, but rather to examine individual differences in cognitive abilities and performance on lineup tasks. The tasks which underlie each set of lineups (as described in the caption of Fig. 2 in the revision) are common tasks in statistical graphics. In addition, the similarity in test protocol between the VST, the tests from the factor-referenced kit of cognitive tests, and the lineup protocol was intentional; by controlling the experimental protocol, we could reduce participant confusion and the corresponding variability. As our goal was not to examine the elements of visualization design experimentally, controlling those elements adheres to good experimental design principles.*

### 2015-06-26 Paper/InfoVis2015Revision/TestingVisualAptitude.Rnw
Remove note, slight wording change

Lineups provide a quantitative measurement of the effectiveness of a particular plot: if participants consistently identify the target plot rather than the randomly-generated distractors, the plot effectively shows the difference between real data and random noise. 
This removes much of the subjectivity from user evaluations of display effectiveness, and the procedure is simple enough that it does not generally require participants to be very familiar with data-based graphics. \hhnote{should we say something here about using charts only for comparisons? We almost do that already. Maybe just drive that message home? }
While previous research~\cite{lowrie2007solving,mayer1994whom} has examined the link between certain types of graphical perception and spatial skills, it is important to identify any additional visual skills participants utilize to complete the lineup task, as well as better understand demographic characteristics (math education, research experience, age, gender) which may impact performance~\cite{humanfactorslineups}.

@@

We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. \newtext{The last three tests are part of the Kit of Factor-Referenced Cognitive Tests~}\cite{educational1963kit, ekstrom1976manual}. \newtext{These tests measure three separate dimensions of visuospatial ability and have been virtually unchanged in use  since the 1960s, and are thus extremely well-studied. }

The VST measures visual search speed~\cite{goldstein1973validity}; \newtext{we considered visual search as the basis of a lineup evaluation and included this test mainly to identify potential problems in the results, if participants were unable to perform this task.} 
Both  card rotation and paper folding tests measure spatial manipulation ability in \newtext{two- and three dimensions}, while the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
\newtext{we hypothesized that} all of these skills are %at least peripherally
recruited during the lineup task, but \newtext{are interested which of these dominate} in predicting performance on the lineup task.

@@

Statistical lineups depend on the ability to search for a signal amid a set of distractors (visual search) and the ability to infer patterns from stimuli (pattern recognition). Depending on the choice of plot shown in the lineup, the task of identifying the most different plot might require additional abilities from participants, e.g.~polar coordinates depend on the ability to mentally rotate stimuli (spatial rotation) and mentally manipulate graphs (spatial rotation and manipulation). By breaking the lineup task down into its components, we  determine which visuospatial factors most strongly correlate with lineup performance, using carefully chosen\newtext{, externally validated,} cognitive tests to assess these aspects of visuospatial ability.

>>>>

Lineups provide a quantitative measurement of the effectiveness of a particular plot: if participants consistently identify the target plot rather than the randomly-generated distractors, the plot effectively shows the difference between real data and random noise. 
This removes much of the subjectivity from user evaluations of display effectiveness, and the procedure is simple enough that it does not generally require participants to be very familiar with data-based graphics. 
While previous research~\cite{lowrie2007solving,mayer1994whom} has examined the link between certain types of graphical perception and spatial skills, it is important to identify any additional visual skills participants utilize to complete the lineup task, as well as better understand demographic characteristics (math education, research experience, age, gender) which may impact performance~\cite{humanfactorslineups}.

@@

We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. The last three tests are part of the Kit of Factor-Referenced Cognitive Tests~\cite{educational1963kit, ekstrom1976manual}. These tests measure three separate dimensions of visuospatial ability and have been virtually unchanged in use  since the 1960s, and are thus extremely well-studied.

The VST measures visual search speed~\cite{goldstein1973validity}; we considered visual search as the basis of a lineup evaluation and included this test mainly to identify potential problems in the results, if participants were unable to perform this task. 
Both  card rotation and paper folding tests measure spatial manipulation ability in two- and three dimensions, while the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
we hypothesized that all of these skills are recruited during the lineup task, but are interested which of these dominate in predicting performance on the lineup task.

@@

Statistical lineups depend on the ability to search for a signal amid a set of distractors (visual search) and the ability to infer patterns from stimuli (pattern recognition). Depending on the choice of plot shown in the lineup, the task of identifying the most different plot might require additional abilities from participants, e.g.~polar coordinates depend on the ability to mentally rotate stimuli (spatial rotation) and mentally manipulate graphs (spatial rotation and manipulation). By breaking the lineup task down into its components, we  determine which visuospatial factors most strongly correlate with lineup performance, using carefully chosen, externally validated, cognitive tests to assess these aspects of visuospatial ability.
